\chapter{The user model of \sysname{}}
\label{chap-user-model}

In order to be an effective user of \sysname{}, you need to understand
its \emph{user model} \index{user model}, i.e., the way the user is
supposed to think about the structure of the score being created.

\section{Composition}
\label{sec-model-composition}

The basic abstraction of \sysname{} is that of a \emph{composition}
\index{composition}.  The composition is the in-memory representation
of an entire piece of music.  \sysname{} can have several
compositions simultaneously in memory.

Loading a \sysname{} file into the editor creates a new composition,
and a composition can be saved either to the file from which it was
initially loaded, or to some other file.  A new composition can be
created (which would then not yet be associated with a file).

The page layout algorithm of \sysname{} has been designed to be able
to cope with very large compositions, while still updating the display
in real time.  Several hundreds of pages should not be a problem.
Though, if you change some global parameter which effects the complete
layout of the entire score, such as the spacing style, and you have a
very large composition, you might have to wait a while before
\sysname{} has completed the calculation of the new page layout.
Ordinary local changes that do not impact page layout elsewhere only
require a very small amount of work for \sysname{}, so updating the
screen should be immediate.

\section{Layer}

A composition can have several \emph{layers}\index{layer}.  The layers
of a composition are temporally superimposed.  You might think of a layer
as corresponding roughly to a \emph{voice} \index{voice} or a
\emph{part} \index{part} of the music.

The difference between a layer and a part (\sysname{} does not know
about parts) is that a layer has a particular \emph{instrument}
\index{instrument} assigned to it, whereas a part can have several
different instruments, for instance when the same musician plays
several instruments (though not simultaneously).  This difference
poses no problem, because it is possible to organize \sysname{} layers
so that their musical material does not overlap in time, even though
layers conceptually do.

\section{Views}

A layers of a composition contain the complete musical material of en
entire piece of music, but they do not contain any information on how
this material is to be presented to the user.  Instead this
information is contained in one or more \texttt{views} that are part
of the composition.

\section{Element}

Each layer is a sequence of \emph{elements}\index{element}.  There are
different kinds of element, \emph{clusters}\index{cluster},
\emph{silences}\index{silence}, and \emph{barlines}\index{barline}.
For the remainder of this section, we ignore the possibility of
an element being a barline, since a barline does not have any
interesting properties such as duration.

In the layer, the first element starts at time zero from the
beginning of the slice, and each subsequent element starts when the
preceding one ends. 

Each element (except the barline) has a particular \emph{duration}
\index{duration}.  The duration is indicated by a \emph{notehead}
\index{notehead} and a \emph{number of dots}. \index{number of dots}
\index{dots!number of|)} Notice that the notehead property exists even
if the element is a silence.  It is used to indicate what kind of
character to use for the silence.

\section{Cluster}

A \emph{cluster}\index{cluster} is roughly equivalent to a chord, in
that it contains all the notes that should be played simultaneously
within a layer.  In addition to the properties inherited from the
element, each cluster has a \emph{stem direction}. \index{stem
direction} \index{direction!of stem|)}

Clusters have zero or more \emph{notes}.  When there are zero notes in
the cluster, it is not displayed at all, but the layout algorithm
takes it into account with its ordinary duration.  This allows the
user to introduce spaces in the music that are not filled with
explicit silence. 

The stem direction indicated in the cluster may or may not be
respected by the layout algorithm.  First, if the stem direction is
\emph{auto}, the layout algorithm will use its knowledge about music
engraving to display the cluster either with the stem up or with the
stem down.  When the stem direction is \emph{down} or \emph{up} as
chosen by the user, the layout algorithm always respects this choice
when the cluster is not part of a beam group.  Currently, all the
stems of a beam group are displayed in the same direction (we will
change this in the future, at least for beam groups where the elements
are on different staves).  Thus, when the cluster is part of a beam
group the stem directions used are determined by the first element of
the group.  When the first element of the group has stem direction
\emph{auto}, then \sysname{} will compute the best stem direction for all
the elements of the group.  When the stem direction of the first
element is either \emph{down} or \emph{up}, \sysname{} will honor that
choice.  Obviously, in this case, stem directions of the other
elements are ignored. 

\section{Silence}

A silence is an other example of an element.  Although always
displayed with the traditional characters, a silence still has a
notehead, which is used in this case to compute the duration of the
silence, and thus what character to use to display it and to determine
spacing. 

\section{Barline}

A barline is another kind of element.  It has no interesting
properties in itself.  Apart from being displayed, it is used to align
musical material in different layers.  Thus, \sysname{} does not bother to
verify that each layer has the same duration between two barlines.
It simply lines up every measure bar in every layer.  If some
layer has a measure that has a shorter duration than the corresponding
measure in a different layer, the shorter one will be padded by empty
space so that they line up. 

Also, \sysname{} does not verify that there is the same number of barlines
in each layer.  When a layer runs out of measure bars, the remaining
ones are lined up without taking into account the layer that ran out.
